Maximizing Ink in Symmetric Partial Edge Drawings of $k$-plane Graphs

### 2018-01-08 eurocg/paper.tex
section 4.1 beta, 4.2 alpha

\subsection{Trees}
\label{sec:tree}
\begin{figure}[tbp]
	\centering
	\includegraphics{tree_illustration}
	\caption{One edge $ u $ with intersecting segments. The shaded areas are the stubs, $ p(u) $ is the parent of $ u $. The stubs in the boxed area contribute to $ \solshort(u) $.}
	\label{fig:tree}
\end{figure}
Here we assume that the intersection graph $ C = (V,E) $ is a rooted tree of maximum degree $ k $. We give a bottom up dynamic programming algorithm for solving \maxsped on $ C $. For a vertex $ u \in V $ let $ p(u) $ denote the parent of $ u $ in $ C $ and let $ c(u) $ denote the set of its children. For a vertex $ u \in V $ with parent $ p(u) \in V $ let $ a \in \mathbb{R}_+ $ be the shorter length between one of $ s(u) $'s endpoint and the intersection point between $ s(u) $ and $ s(p(u)) $. Now we can define the following two values, allowing us to categorize the stub-lengths into those not affecting the parents choice of stub and those which don't:
%Since we know $ C $ is a tree every vertex $ u \in V $ with degree $ \delta $ has a parent $ p(u) \in V $. For this parent there exists one stub length $ \ell_j(p(u)) $ in the sequence of stub lengths $ \ell_0(p(u)),\dots,\ell_{\delta}(p(u)) $ such that drawing $ p(u) $ with these stubs does not intersect a completely drawn $ u $, while drawing $ p(u) $ with $ \ell_{j + 1}(p(u)) $, i.e. the next longer stub length, intersects $ \ell_0(u) $. We call the solution which would intersect the parent $ \sollong(u) $ and the one not intersecting the parent $ \solshort(u) $. See Figure~\ref{fig:tree} for an illustration. More formally let $ j $ be the index such that stubs of length $ \ell_j(p(u)) $ do not intersect $ s(u) $, but $ \ell_{j +1}(p(u)) $ would, we set:
\begin{align*}
	\solshort(u) = \max\{T_1(u),\dots,T_a(u)\} &&
	\sollong(u) = \max\{T_0(u),\dots,T_{\delta}(u)\}
\end{align*}
It remains to define $ T_i(u) $ for a vertex $ u $ and stub length $ \ell_i(u) $:
\begin{align}
	\label{rec:tree}
	T_i(u) = \ell_i(u) + \sum_{v \in c(u)}
	\begin{cases}
		\solshort(v) & \text{if } s(u) \text{ with length } \ell_i \text{ intersects } s(v) \\
		\sollong(v) & \text{else}
	\end{cases}
\end{align}
Solving Recurrence~\ref{rec:tree} yields an optimal solution to the \maxsped problem on $ G $ with drawing $ \Gamma $ as the following lemma shows.

\begin{lemma}
	\label{lem:tree_correctness}	
	Let $ G $ be a simple graph and $ \Gamma $ a straight-line drawing of $ G $. If the intersection graph $ C = (V,E) $ of $ G $ is a tree with root $ r \in V $ and maximum degree $ k $, then $ \sollong(r) = \max \{T_i(r) \mid 0 \leq i \leq k \} $ is an optimal solution to the \maxsped problem on $ G $ with drawing $ \Gamma $.
\end{lemma}
\begin{proof}
	Let $ G $, $ \Gamma $ and $ C = (V,E) $ as above and $ u \in V $ a leaf of $ C $. Since $ c(u) $ is empty we simply get $ \solshort(u) = T_0(u) = \ell_1(u) $ and $ \sollong(u) = T_1(u) = \max\{\ell_0(u), \ell_1(u)\} $. Now let $ u \in V $ with degree $ \delta $ be an inner vertex. The values $ \solshort(v) $ and $ \sollong(v) $ are computed for all $ v \in c(u) $. $ T_i(u) $ for one $ 0 \leq i \leq \delta $ is then the sum of the maximum stubs we can draw among the children plus the stub length $ \ell_i(u) $ itself, which are just the values we are interested in. Setting then $ \sollong(u) $ and $ \solshort(u) $ as above gives the two necessary values for $ p(u) $.
%	and $ t = \sollong(r) $ is the maximum value for some $ T_i(r) $'s and $ r $ the root. Now let $ t' $ be the value of a solution to the \maxsped problem on $ G $ with drawing $ \Gamma $ and $ t' > t $. This solution has to be the sum of a set of stub lengths for each edge in $ G $ such that no two stubs intersect by definition. Further we know the stubs can be categorized into two sets relative to their parent. Recurrence~\ref{rec:tree} builds just the maximum over the $ \solshort(v) $ and $ \sollong(v) $ values for all $ v \in c(u) $, such that no two stubs intersect. Hence if $ t $ was not optimal we would need to find at least a third stub length for at least one child $ v \in c(u) $ which was neither equal to $ \sollong(v) $ nor $ \solshort(v) $.
\end{proof}

@@

\begin{lemma}
	\label{lem:compute_vertex}
	Let $ C = (V,E) $ be a tree with maximum degree $ k $ and vertex weights $ \ell_0,\dots,\ell_{\delta} $ for every $ u \in V $ with degree $ \delta $. We can compute $ \solshort(u) $ and $ \sollong(u) $ in $ O(\delta) $ time and space, if the values $ \sollong(v) $ and $ \solshort(v) $ are computed for all children $ v \in c(u) $.
\end{lemma}
\begin{proof}
	 Let $ u \in V $ be a vertex with degree $ \delta < k $ for which we have computed $ \sollong $ and $ \solshort $ for every of its children. The crucial part is that $ T_i(u) $ can be built iteratively by going through the list of stub lengths for $ u $ in ascending order. First we compute $ T_1(u) $, this is the solution for $ u $ in which we are able to draw the children with any of their stub lengths. This means we need to build the sum of $ \ell_1(u) $ and the $ \sollong(v) $ for every $ v \in c(u) $ which runs in $ O(\delta) $. The same can be done for $ T_0(u) $, but taking all the $ \solshort(v) $ values for the children $ v \in c(u) $. Assume we computed $ T_j(u) $ for all stub lengths $ \ell_i(u) $ with $ i \leq j $. Now we want to compute $ T_{j+1}(u) $ (see Figure~\ref{fig:tree}). By drawing $ s(u) $ with stub length $ \ell_{j +1} $ we make it a bit longer since we consider the $ \ell_i(u) $ in ascending order. Hence by doing so we now potentially intersect the corresponding edge of the $ j $-th child $ v_j \in c(u) $. Since $ s(v_j) $ is the only segment affected by drawing $ u $ with stub length $ \ell_{j + 1} $ instead of $ \ell_j $ we can compute $ T_{j+1}(u) $ in $O(1)$ time by simply subtracting $ \sollong(v) $ and adding $ \solshort(v) $.

@@

\begin{theorem}
	\label{thm:tree}
	Let $ G $ be a simple graph with $m$ edges and $ \Gamma $ a straight-line drawing of $ G $. If the intersection graph $ C = (V,E) $ of $ G $ is a tree with maximum degree $ k \in \mathbb{N} $ the \maxsped problem can be solved in $ O(mk + m \log m)) $ time and space.
\end{theorem}
\begin{proof}
	Since $ C $ is known to be a tree, and hence $ |E| = O(m) $, we can compute the intersection graph and the stub lengths in $ O(m\log m) $ time and space.

	To solve the \maxsped problem on $ G $ we need to compute $ \max \{T_i(r) \mid 0 \leq i \leq k \} $ for the root $ r \in V $ of $ C $ by Lemma~\ref{lem:tree_correctness}. Doing this bottom up requires one pass through $ C $ hence by applying Lemma~\ref{lem:compute_vertex} we need $ O(mk) $ time to compute all $ T_i(u) $ for all $ u \in V $. For every vertex we only need to compute $ O(k) $ values and store two of them, namely $ \sollong(u) $ and $ \solshort(u) $, so we do not need more than $ O(mk) $ memory. Adding the bounds yields the desired runtime of $ O(mk + m \log m)) $.

	Retrieving the actual stub length for each edge in $ G $ needed to produce the SPED can be done by backtracking the subsolutions which requires at most $ O(mk) $ time.
\end{proof}

>>>>

\subsection{Trees}
\label{sec:tree}
\begin{figure}[tbp]
	\centering
%	\includegraphics{tree_illustration}
	\caption{One edge $ u $ with intersecting segments. The shaded areas are the stubs, $ p(u) $ is the parent of $ u $. The stubs in the boxed area contribute to $ \solshort(u) $.}
	\label{fig:tree}
\end{figure}
Here we assume that the intersection graph $ C = (V,E) $ is a rooted tree of maximum degree $ k $. We give a bottom up dynamic programming algorithm for solving \maxsped on $ C $. For a vertex $ u \in V $ let $ p(u) $ denote the parent of $ u $ in $ C $ and let $ c(u) $ denote the set of its children. For a vertex $ u \in V $ with parent $ p(u) \in V $ let $ a \in \mathbb{R}_+ $ be the shorter length between one of $ s(u) $'s endpoint and the intersection point between $ s(u) $ and $ s(p(u)) $. Now we can define the following two values, allowing us to categorize the stub-lengths into those not affecting the parents choice of stub and those which don't:
%Since we know $ C $ is a tree every vertex $ u \in V $ with degree $ \delta $ has a parent $ p(u) \in V $. For this parent there exists one stub length $ \ell_j(p(u)) $ in the sequence of stub lengths $ \ell_0(p(u)),\dots,\ell_{\delta}(p(u)) $ such that drawing $ p(u) $ with these stubs does not intersect a completely drawn $ u $, while drawing $ p(u) $ with $ \ell_{j + 1}(p(u)) $, i.e. the next longer stub length, intersects $ \ell_0(u) $. We call the solution which would intersect the parent $ \sollong(u) $ and the one not intersecting the parent $ \solshort(u) $. See Figure~\ref{fig:tree} for an illustration. More formally let $ j $ be the index such that stubs of length $ \ell_j(p(u)) $ do not intersect $ s(u) $, but $ \ell_{j +1}(p(u)) $ would, we set:
\begin{align*}
	\solshort(u) = \max\{T_1(u),\dots,T_a(u)\} &&
	\sollong(u) = \max\{T_0(u),\dots,T_{\delta}(u)\}
\end{align*}
It remains to define $ T_i(u) $ for a vertex $ u $ and stub length $ \ell_i(u) $:
\begin{align}
	\label{rec:tree}
	T_i(u) = \ell_i(u) + \sum_{v \in c(u)}
	\begin{cases}
		\solshort(v) & \text{if } s(u) \text{ with length } \ell_i \text{ intersects } s(v) \\
		\sollong(v) & \text{else}
	\end{cases}
\end{align}
By solving Recurrence~\ref{rec:tree} we find an optimal solution to the \maxsped problem on $ G $ with drawing $ \Gamma $.	Let $ C = (V,E) $ as above and $ u \in V $ a leaf of $ C $. Since $ c(u) $ is empty we simply get $ \solshort(u) = T_0(u) = \ell_1(u) $ and $ \sollong(u) = T_1(u) = \max\{\ell_0(u), \ell_1(u)\} $. Now let $ u \in V $ with degree $ \delta $ be an inner vertex. The values $ \solshort(v) $ and $ \sollong(v) $ are computed for all $ v \in c(u) $. $ T_i(u) $ for one $ 0 \leq i \leq \delta $ is then the sum of the maximum stubs we can draw among the children plus the stub length $ \ell_i(u) $ itself, which are just the values we are interested in. Setting then $ \sollong(u) $ and $ \solshort(u) $ as above gives the two necessary values for $ p(u) $.
%\begin{lemma}
%	\label{lem:tree_correctness}	
%	Let $ G $ be a simple graph and $ \Gamma $ a straight-line drawing of $ G $. If the intersection graph $ C = (V,E) $ of $ G $ is a tree with root $ r \in V $ and maximum degree $ k $, then $ \sollong(r) = \max \{T_i(r) \mid 0 \leq i \leq k \} $ is an optimal solution to the \maxsped problem on $ G $ with drawing $ \Gamma $.
%\end{lemma}
%\begin{proof}
%	Let $ G $, $ \Gamma $ and $ C = (V,E) $ as above and $ u \in V $ a leaf of $ C $. Since $ c(u) $ is empty we simply get $ \solshort(u) = T_0(u) = \ell_1(u) $ and $ \sollong(u) = T_1(u) = \max\{\ell_0(u), \ell_1(u)\} $. Now let $ u \in V $ with degree $ \delta $ be an inner vertex. The values $ \solshort(v) $ and $ \sollong(v) $ are computed for all $ v \in c(u) $. $ T_i(u) $ for one $ 0 \leq i \leq \delta $ is then the sum of the maximum stubs we can draw among the children plus the stub length $ \ell_i(u) $ itself, which are just the values we are interested in. Setting then $ \sollong(u) $ and $ \solshort(u) $ as above gives the two necessary values for $ p(u) $.
%	and $ t = \sollong(r) $ is the maximum value for some $ T_i(r) $'s and $ r $ the root. Now let $ t' $ be the value of a solution to the \maxsped problem on $ G $ with drawing $ \Gamma $ and $ t' > t $. This solution has to be the sum of a set of stub lengths for each edge in $ G $ such that no two stubs intersect by definition. Further we know the stubs can be categorized into two sets relative to their parent. Recurrence~\ref{rec:tree} builds just the maximum over the $ \solshort(v) $ and $ \sollong(v) $ values for all $ v \in c(u) $, such that no two stubs intersect. Hence if $ t $ was not optimal we would need to find at least a third stub length for at least one child $ v \in c(u) $ which was neither equal to $ \sollong(v) $ nor $ \solshort(v) $.
%\end{proof}

@@

\begin{lemma}
	\label{lem:compute_vertex}
	Let $ C = (V,E) $ be a tree with maximum degree $ k $ and vertex weights $ \ell_0,\dots,\ell_{\delta} $ for every $ u \in V $ with degree $ \delta $. We can compute $ \solshort(u) $ and $ \sollong(u) $ in $ O(\delta) $ time and space, if the values $ \sollong(v) $ and $ \solshort(v) $ are computed for all children $ v \in c(u) $.
\end{lemma}
\begin{proof}
	 Let $ u \in V $ be a vertex with degree $ \delta < k $ for which we have computed $ \sollong $ and $ \solshort $ for every of its children. 
%	 The crucial part is that $ T_i(u) $ can be built iteratively by going through the list of stub lengths for $ u $ in ascending order. 
	First we compute $ T_0(u) $ and $ T_1(u) $, this can be done since they represent the longest (whole edge) and shortest stub lengths. 	 
	 Now assume we computed $ T_j(u) $ for all stub lengths $ \ell_i(u) $ with $ i \leq j $, hence we want to compute $ T_{j+1}(u) $ next. Doing so only requires to consider the next stub-length $ \ell_{j + 1}(u) $, since all other affected by drawing $ s(u) $ with $ \ell_{j+1}(u) $ where already considered. Hence we can compute $ T_{j+1}(u) $ in $O(1)$ time by simply subtracting $ \sollong(v) $ and adding $ \solshort(v) $.

@@

\begin{theorem}
	\label{thm:tree}
	Let $ G $ be a simple graph with $m$ edges and $ \Gamma $ a straight-line drawing of $ G $. If the intersection graph $ C = (V,E) $ of $ G $ is a tree with maximum degree $ k \in \mathbb{N} $ the \maxsped problem can be solved in $ O(mk + m \log m)) $ time and space.
\end{theorem}
\begin{proof}
	Since $ C $ is known to be a tree, and hence $ |E| = O(m) $, we can compute the intersection graph and the stub lengths in $ O(m\log m) $ time and space.
	With Lemma~\ref{lem:compute_vertex} and Recurrence~\ref{rec:tree} we need one pass through the tree to find the roots maximum in $ O(mk) $ time. Adding the bounds yields the desired runtime of $ O(mk + m \log m)) $.
%	To solve the \maxsped problem on $ G $ we need to compute $ \max \{T_i(r) \mid 0 \leq i \leq k \} $ for the root $ r \in V $ of $ C $ as seen above. Doing this bottom up requires one pass through $ C $ hence by applying Lemma~\ref{lem:compute_vertex} we need $ O(mk) $ time to compute all $ T_i(u) $ for all $ u \in V $. For every vertex we only need to compute $ O(k) $ values and store two of them, namely $ \sollong(u) $ and $ \solshort(u) $, so we do not need more than $ O(mk) $ memory. Adding the bounds yields the desired runtime of $ O(mk + m \log m)) $.
%	
	Retrieving the actual stub length for each edge in $ G $ needed to produce the SPED can be done by backtracking the subsolutions which requires at most $ O(mk) $ time.
\end{proof}
